Software architectures for workflow integration in science gateways

### 2016-04-21 paper.tex
Changed color of notes
Updates in Section 1

\newcommand{\todo}[1]{\color{blue}\xspace[\emph{#1}]\xspace\color{black}}
\newcommand{\note}[1]{\color{green}\textbf{Note: }\textit{#1} \color{black}}

@@

\begin{abstract}
%% Text of abstract
  We investigate software architectures to integrate workflow engines
  in science gateways. Based on our experience with the development
  and operations of the CBRAIN and VIP science gateways, and our past
  involvement in the SHIWA project, we abstract the software
  components and interactions in 6 architectures that we compare using
  metrics measuring integration effort, robustness, extensibility,
  scalability and other specific features. \todo{a word on results}
\end{abstract}

@@

The remainder of this Section provides background information and
definitions about workflow engines and science gateways. In
Section~\ref{sec:architectures}, we describe 6 architectures in a
consistent framework based on the functional interactions between the
main software components involved. In Section~\ref{sec:evaluation}, we
introduce metrics to measure integration effort, robustness,
extensibility, scalability and other specific features related to the
architectures. We use these metrics to evaluate and compare the 6
architectures. The paper closes on a discussion and conclusion that
highlight architectural recommendations for science gateway and
workflow engine developers.

@@

In the last decade, the e-Science workflow community has developed
workflow systems to help developers access distributed infrastructures
such as clusters, grids, clouds and web services. These efforts
resulted in tools such as Askalon~\cite{fahringer2005askalon},
Hyperflow~\cite{balis2016hyperflow}, MOTEUR~\cite{GLAT-08i},
Pegasus~\cite{deelman2005pegasus,Deelman201517},
Swift~\cite{zhao2007swift}, Taverna~\cite{oinn2004taverna},
Triana~\cite{taylor2007triana}, VisTrails~\cite{callahan2006managing},
WS-PGRADE/gUSE~\cite{Kacsuk2012}\footnote{\todo{WS-PGRADE/gUSE is the
    science gateway, check what the engine is called now}}, etc. Such
workflow engines usually describe applications in a specific language
that offers parallel data and control flow constructs, visual edition
features, links with domain-specific tool repositories, provenance
recording, among other features.
% Descriptions and experiments
%conducted with e-Science workflow engines were published in journals
%such as Future Generation Computer Systems and conference venues such
%as WORKS.

At the same time, specific toolboxes were emerging in various
scientific domains to facilitate the assembly of software components
in consistent pipelines. In neuroimaging, our primary domain of
interest, tools such as Nipype~\cite{gorgolewski2011nipype},
PSOM~\cite{bellec2012pipeline}, SPM~\cite{ashburner2011spm} and
FSL~\cite{Jenkinson2012782} provide abstractions and functions to
handle the data and computing flow between processes implemented in a
variety of programming languages. Such tools were interfaced to
computing systems, in particular clusters, to create tasks, handle
their dependencies, and execute them. For instance, FSL can launch
tasks on SGE through its \texttt{fsl\_sub} tool, Nipype has execution
plugins for S/OGE, Torque, IPython, ssh and multi-processor servers,
and PSOM is interfaced to Torque, SGE, MOAB, LSF and Condor. Some of
these tools also support advanced features such as provenance tracking
(Nipype, PSOM), redundancy detection across analyses to avoid
re-computation (PSOM), and so on. A wide-array of workflows have been
implemented using these pipelines and are now widely shared across
neuroimaging groups world-wide. \todo{Provide more context on that?}
These domain-specific engines represent a tremendous opportunity for
science gateways to leverage existing tools and applications. They
nicely complement e-Science engines that are more oriented towards the
exploitation of distributed computing infrastructures, in particular
grids and clouds.

For the sake of the analysis conducted in this paper, we
define a \emph{workflow engine} (also abbreviated \emph{engine}) as a
software that submits interdependent computing \emph{tasks} to an
infrastructure (local server, cluster, grid or cloud) based on a
workflow description (a.k.a \emph{workflow}) and using input data that
may be files, data base entries, or simple parameter values. Although
simplistic, this definition covers both e-Science and domain-specific
engines. Some workflow engines, usually e-Science ones, may transfer
data across the infrastructure, and others, usually domain-specific
ones, may leave it to external processes. Workflows may be expressed
in any language, including high-level XML or JSON dialects such as
\texttt{Scufl} or \texttt{Hyperflow}, and low-level scripts such as
\texttt{bash}.  \todo{Mention other workflow definitions,
  e.g. abstract/concrete, etc?} \todo{Use ttt everywhere for technical
  terms (ssh, etc)?}

\todo{The infrastructure consists of the computing and storage resources
involved in the workflow execution, and the software services used to
access these resources. Infrastructures can be servers, clusters,
grids or clouds. Some workflow engines may assume specific
characteristics about the infrastructure, such as the presence of a
shared file system between the computing nodes.
}

@@

Scalability is also an important concern, since science gateways are
multi-user systems, manage substantial amounts of data and execute
computationally-intensive analyses. For this purpose, science gateways
may include different instances of the same engine type and implement
some load-balancing policy among them. They may also enforce
scheduling policies on the infrastructure, e.g. to ensure fairness
among users~\cite{FERR-14} or to improve
fault-tolerance~\cite{FERR-13}. \todo{Remove these self-citations or
  cite more works on this topic.}

@@

Various science gateways have been developed in the last years,
including science gateway frameworks used as building blocks to
assemble science gateways. The main frameworks are
WS-PGRADE/gUSE~\cite{Kacsuk2012}, the Catania Science Gateway
Framework~\cite{Ardizzone2012} and Apache
Airavata~\cite{marru2011apache} \todo{According to
  https://airavata.apache.org, Genapp is based on Airavata, we could
  refer to it somewhere}. Numerous science gateway instances were
built using such frameworks, in various disciplines. Other gateway
instances were also built independently from any framework, using
regular web development tools and various libraries to manage
distributed systems.

\todo{Transition needed}
CBRAIN\footnote{\url{http://cbrain.mcgill.ca}}~\cite{SHER-14} is an
open-source, web-based collaborative platform addressing the
challenges of data-heavy, computation-intensive neuroimaging
research. It offers transparent web-based access to remote data
sources, distributed computing sites, and an array of processing and
visualization tools, within a secure environment. CBRAIN is deployed
at a wide range of computing sites, including nine national HPC
centers in Canada, one in Korea, one in Germany, and on numerous local
servers. As September 2015, CBRAIN has served over 370 users across 50
cities in 19 countries who produced several high-impact results in the
field of neurosciences.

@@

\todo{SHIWA Simulation Platform.}

@@

The CBRAIN-FSL integration is also remarkable as it allows to leverage
FSL pipelines that are written in a very low-level workflow language
(Linux executables and script) that submits tasks uniformly through a
specific tool called \texttt{fsl\_sub}. To integrate FSL in CBRAIN,
\texttt{fsl\_sub} was modified to implement interaction
\texttt{e}\footnote{See \texttt{fsl\_sub} modified script available at
  \url{https://github.com/aces/cbrain-plugins-neuro}} \todo{Add a
  diagram and maybe a code excerpt to illustrate that.}.

@@

The particular case of interaction \texttt{g} (workflow language
conversion) should be further discussed since it is very likely that
this interaction could not be easily generalized to any workflow
language. For instance, converting FSL, PSOM or Nipype pipelines to
any other workflow language is very tricky because these engines rely
on general-purpose programming languages (bash, Octave, Python, etc)
which are much richer than scientific workflow languages.

>>>>

\definecolor{ao(english)}{rgb}{0.0, 0.5, 0.0}
\newcommand{\todo}[1]{\color{blue}\xspace[\emph{#1}]\xspace\color{black}}
\newcommand{\note}[1]{\color{ao(english)}\textbf{Note: }\textit{#1} \color{black}}

@@

\begin{abstract}
%% Text of abstract
  We investigate software architectures to integrate workflow engines
  in science gateways. Based on our experience with the development
  and operations of the CBRAIN and VIP science gateways, and our past
  involvement in the SHIWA project, we abstract the software
  components and interactions in 6 architectures that we compare using
  metrics measuring integration effort, robustness, extensibility,
  scalability and other specific features. Results show the pros and
  cons of the different architectures. Tight integration and
  sub-tasking require the least amount of integration effort and are
  also the most robust. Most architectures perform the same in terms
  of extensibility. The pool model is the most
  scalable. Meta-workflows are only available in nested workflows and
  workflow import.
\end{abstract}

@@

The remainder of this Section provides background information and
definitions about workflow engines, science gateways and
infrastructures. In Section~\ref{sec:architectures}, we describe 6
architectures in a consistent framework based on the functional
interactions between the main software components involved. In
Section~\ref{sec:evaluation}, we introduce metrics to measure
integration effort, robustness, extensibility, scalability and other
specific features related to the architectures. We use these metrics
to evaluate and compare the 6 architectures. The paper closes on a
discussion and conclusion that highlight architectural recommendations
for science gateway and workflow engine developers.

@@

In the last decade, the e-Science workflow community has developed
workflow systems to help developers access distributed infrastructures
such as clusters, grids, clouds and web services. These efforts
resulted in tools such as Askalon~\cite{fahringer2005askalon},
Hyperflow~\cite{balis2016hyperflow}, MOTEUR~\cite{GLAT-08i},
Pegasus~\cite{deelman2005pegasus,Deelman201517},
Swift~\cite{zhao2007swift}, Taverna~\cite{oinn2004taverna},
Triana~\cite{taylor2007triana}, VisTrails~\cite{callahan2006managing},
WS-PGRADE~\cite{Kacsuk2012}, etc. Such
workflow engines usually describe applications in a specific language
that offers parallel data and control flow constructs, visual edition
features, links with domain-specific tool repositories, provenance
recording, among other features.
% Descriptions and experiments
%conducted with e-Science workflow engines were published in journals
%such as Future Generation Computer Systems and conference venues such
%as WORKS.

At the same time, specific toolboxes were emerging in various
scientific domains to facilitate the assembly of software components
in consistent pipelines. In neuroimaging, our primary domain of
interest, tools such as Nipype~\cite{gorgolewski2011nipype},
PSOM~\cite{bellec2012pipeline}, SPM~\cite{ashburner2011spm} and
FSL~\cite{Jenkinson2012782} provide abstractions and functions to
handle the data and computing flow between processes implemented in a
variety of programming languages. Such tools were interfaced to
computing systems, in particular clusters, to create tasks, handle
their dependencies, and execute them. For instance, FSL can launch
tasks on SGE through its fsl\_sub tool, Nipype has execution
plugins for S/OGE, Torque, IPython, ssh and multi-processor servers,
and PSOM is interfaced to Torque, SGE, MOAB, LSF and Condor. Some of
these tools also support advanced features such as provenance tracking
(Nipype, PSOM), redundancy detection across analyses to avoid
re-computation (PSOM), and so on. A wide-array of workflows have been
implemented using these pipelines and are now widely shared across
neuroimaging groups world-wide. \todo{Provide more context on that?}
These domain-specific engines represent a tremendous opportunity for
science gateways to leverage existing tools and applications. They
nicely complement e-Science engines that are more oriented towards the
exploitation of distributed computing infrastructures, in particular
grids and clouds.

For the sake of the analysis conducted in this paper, we
define a \emph{workflow engine} (also abbreviated \emph{engine}) as a
software that submits interdependent computing \emph{tasks} to an
infrastructure (local server, cluster, grid or cloud) based on a
workflow description (a.k.a \emph{workflow}) and using input data that
may be files, data base entries, or simple parameter values. Although
simplistic, this definition covers both e-Science and domain-specific
engines. Some workflow engines, usually e-Science ones, may transfer
data across the infrastructure, and others, usually domain-specific
ones, may leave it to external processes. Workflows may be expressed
in any language, including high-level XML or JSON dialects such as
Scufl or Hyperflow, and low-level scripts such as
Bash.

@@

Scalability is also an important concern, since science gateways are
multi-user systems, manage substantial amounts of data and execute
computationally-intensive analyses. For this purpose, science gateways
may include different instances of the same engine type and implement
some load-balancing policy among them. They may also enforce
scheduling policies on the infrastructure, e.g. to ensure fairness
among users~\cite{FERR-14,CPE:CPE3708} or to improve
fault-tolerance~\cite{wilkins2008teragrid,FERR-13}.

@@

Various science gateways have been developed in the last years,
including science gateway frameworks used as building blocks to
assemble science gateways. The main frameworks are
WS-PGRADE/gUSE~\cite{Kacsuk2012}, the Catania Science Gateway
Framework~\cite{Ardizzone2012} and Apache
Airavata~\cite{marru2011apache}. Numerous science gateway instances
were built using such frameworks, in various disciplines. Other
gateway instances were also built independently from any framework,
using regular web development tools and various libraries to manage
distributed systems. CBRAIN and VIP are two of such science gateways
that were initially developed for medical imaging analysis and from
which we gathered experience about the architectures described in
this paper.

CBRAIN\footnote{\url{http://cbrain.mcgill.ca}}~\cite{SHER-14} is an
open-source, web-based collaborative platform addressing the
challenges of data-heavy, computation-intensive neuroimaging
research. It offers transparent web-based access to remote data
sources, distributed computing sites, and an array of processing and
visualization tools, within a secure environment. CBRAIN is deployed
at a wide range of computing sites, including nine national HPC
centers in Canada, one in Korea, one in Germany, and on numerous local
servers. As September 2015, CBRAIN has served over 370 users across 50
cities in 19 countries who produced several high-impact results in the
field of neurosciences.

@@

The architectures described hereafter also build on experience with
the SHIWA Simulation Platform~\cite{terstyanszky2014enabling}. The
SHIWA Simulation Platform is a WS-PGRADE/gUSE science gateway that was
developed in European projects
SHIWA\footnote{\url{http://www.shiwa-workflow.eu}} and
ER-flow\footnote{\url{http://www.erflow.eu}}. It focuses on workflow
interoperability for various disciplines, in particular through the
building of meta-workflows, i.e. workflows that are built out of other
workflows. It implements the so-called Coarse-Grained Interoperability
framework where workflow engines are integrated in the science gateway
as other applications and invoked through a dedicated service. The
Fine-Grained Interoperability framework was also developed in SHIWA,
where conversions between workflow languages are achieved through a
common intermediate
representation~\cite{plankensteiner-prodan-etal:2013}.

\subsection{Infrastructures}

In this paper, infrastructure consists of the computing and storage
resources involved in the workflow execution, and the software
services used to access these resources. Infrastructures can be
servers, clusters, grids or clouds. Some workflow engines and science
gateways may assume specific characteristics about the infrastructure,
such as the presence of a shared file system between the computing
nodes, the availability of a global task meta-scheduler, the presence
of a file catalog, etc. In the analysis presented below, such specific
functions are not discussed. Instead, we consider the infrastructure
as an abstract system that can execute tasks and store files
regardless of the enabling mechanisms.

@@

The CBRAIN-FSL integration is also remarkable as it allows to leverage
FSL pipelines that are written in a very low-level workflow language
(Linux executables and script) that submits tasks uniformly through a
specific tool called fsl\_sub. To integrate FSL in CBRAIN,
fsl\_sub was modified to implement interaction
\texttt{e}\footnote{See modified fsl\_sub script available at
  \url{https://github.com/aces/cbrain-plugins-neuro}} \todo{Add a
  diagram and maybe a code excerpt to illustrate that.}.

@@

The particular case of interaction \texttt{g} (workflow language
conversion) should be further discussed since it is very likely that
this interaction could not be easily generalized to any workflow
language. For instance, converting FSL, PSOM or Nipype pipelines to
any other workflow language is very tricky because these engines rely
on general-purpose programming languages (Bash, Octave, Python, etc)
which are much richer than scientific workflow languages.
